---
title: "ColorSpace exploration"
format: typst
---

# ============================================================
# Leaf CIELAB color analysis
#
# Processing pipeline for CIELAB histograms exported from
# Fiji/ImageJ leaf segmentation.
#
# Workflow:
#   1. Import RGB/Lab histogram data and Lab ranges
#   2. Add experimental design metadata
#   3. Select the maximum-numbered leaf for each image
#   4. Normalize pixel histograms to proportions
#   5. Convert 8-bit histogram bins to original Lab values
#   6. Compute weighted mean L*, a*, and b* values
#   7. Calculate pairwise CIEDE2000 color differences
#   8. Generate plots and summary tables
#
# Input files:
#   - all_Lab_bins.csv : Lab histogram counts exported by Fiji
#   - Lab_ranges.csv   : original 32-bit Lab ranges for each histogram
#   - design.csv       : experimental design and treatment metadata
#
# Output:
#   - Results/Plots/ : CIEDE2000 plots
#   - Results/Tables/ : CIELAB summary tables
# ============================================================

```{r}
# Librairies
# CRAN packages
pkgs <- c(
  "tibble", "dplyr", "stringr", "purrr", "ggplot2", "patchwork", "car", "multcomp", "multcompView", "emmeans", "ggpubr", "broom", "tidyr", "here", "farver", "ggtext", "ltc", "typstable"
)

# Bioconductor packages
# pkgs_bioc <- c(
#     ""
# )

# Installer les packages CRAN si nécessaire
for (pkg in pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  suppressPackageStartupMessages(
    library(pkg, character.only = TRUE)
    )
}

# Installer les packages Bioconductor si nécessaire
# if (!requireNamespace("BiocManager", quietly = TRUE)) {
#   install.packages("BiocManager")
# }

# Installer les Bioconductor packages
# BiocManager::install(pkgs_bioc, update = FALSE, ask = FALSE)

# Charger les Bioconductor packages
# for (pkg in pkgs_bioc) {
#   suppressPackageStartupMessages(
#     library(pkg, character.only = TRUE)
#   )
# }

# if (!require('name_github_pkg')) remotes::install_github('repo/pkg')
# library('pkg_name')

# WD 
wd <- function(path = NULL) {
  
  if (!requireNamespace("here", quietly = TRUE)) {
    stop("Package 'here' is required. Install it with install.packages('here').")
  }
  
  if (!is.null(path)) {
    
    path <- path.expand(path)
    
    if (dir.exists(path)) {
      
      # définit le projet racine implicitement si possible
      setwd(normalizePath(path))
      message("Using path: ", normalizePath(path))
      return(invisible(path))
      
    } else {
      warning("Path does not exist: ", path)
    }
  }
  
  choice <- readline(
    "Directory not found or not provided. Type 'f' (choose folder) or 'c' (current project root): "
  )
  
  choice <- tolower(choice)
  
  if (choice == "f") {
    
    new_path <- dirname(file.choose())
    
    setwd(normalizePath(new_path))
    message("Selected folder: ", new_path)
    return(invisible(new_path))
    
  } else if (choice == "c") {
    
    root <- here::here()
    setwd(normalizePath(root))
    message("Using here() project root: ", root)
    return(invisible(root))
    
  } else {
    stop("Invalid option")
  }
}

wd("~/Documents/R/memoire/lab_exploration//")

#=== Colour vision accessibility check
#ltc_cvd(ploen)                     # normal + deuteranopia / protanopia / tritanopia
#ltc_cvd("ploen", severity = 0.6) # milder simulation

#ltc_cvd(alger)                     # normal + deuteranopia / protanopia / tritanopia
#ltc_cvd("alger", severity = 0.6) # milder simulation
```

```{bash}
#!/bin/bash

ROOT="/home/seeker/Documents/Mémoire/pictures/exp2_leaves_scan"
OUT="all_Lab_bins.csv"

# header
echo "Batch,Image,Leaf,Channel,Bin,Count" > $OUT

for batch in batch1 batch2 batch3 batch4
do
    find "$ROOT/$batch/croped/color_corrected" \
    -type f \
    -name "*_Lab_*.csv" | while read file

    do
        # fetch metadata
        image=$(basename $(dirname $(dirname "$file")))
        leaf=$(basename $(dirname "$file"))
        batch_name=$(basename "$batch")

        # fetch which L*, a*, b* channel
        channel=$(basename "$file" | \
                  sed -E 's/.*_Lab_([Lab])\.csv/\1/')
        # remove header and print metadata
        tail -n +2 "$file" | \
        awk -F',' \
        -v b="$batch_name" \
        -v i="$image" \
        -v l="$leaf" \
        -v c="$channel" \
        '{
          print b","i","l","c","$1","$2
        }' >> $OUT
    done
done

OUT="Lab_ranges.csv"

# header
echo "Batch,Image,Leaf,Channel,Header" > $OUT

for batch in batch1 batch2 batch3 batch4
do
    find "$ROOT/$batch/croped/color_corrected" \
    -type f \
    -name "*_Lab_*.csv" | while read file
    do
        # fetch metadata
        image=$(basename $(dirname $(dirname "$file")))
        leaf=$(basename $(dirname "$file"))
        batch_name=$(basename "$batch")

        # fetch which L*, a*, b* channel
        channel=$(basename "$file" | \
                  sed -E 's/.*_Lab_([Lab])\.csv/\1/')

        # fetch first line
        header=$(head -n 1 "$file")

        # remove "," ";" from header
        header=$(echo "$header" | tr ',' ';')
        
        echo "$batch_name,$image,$leaf,$channel,\"$header\"" >> $OUT
    done
done
```

```{r}         
dir.create("Results", showWarnings = FALSE)
dir.create("Results/Tables", showWarnings = FALSE)
dir.create("Results/Plots", showWarnings = FALSE)
    
plot_histogram <- function(
    data,
    x = "Bin",
    y = "Count",
    normalize = TRUE,
    treatment_colors = NULL,
    facet_var = NULL,
    facet_direction = c("rows", "cols"),
    treatment_labels = NULL,
    level_labels = NULL,
    average = TRUE,
    individuals = TRUE,
    individual_alpha = 0.3,
    average_linewidth = 1.5,
    xlabel = "CIELAB bin"
){
  # --- Create Interaction if not already present ---
  if (!"Interaction" %in% names(data)) {
  data <- data %>%
    mutate(Interaction = paste(MicB, MicP, sep=":"))
  }

  data <- data %>%
    mutate(
      Interaction = as.character(Interaction)
    )

  facet_direction <- match.arg(facet_direction)

  # --- Optional normalization ---
  if(normalize){
    data <- data %>%
      group_by(ID, Interaction) %>%
      mutate(Prop = .data[[y]] / sum(.data[[y]])) %>%
      ungroup()

    y <- "Prop"
    ylabel <- "Proportion"

  }else{
    ylabel <- y
  }

  # --- Markdown helpers ---
  escape_md <- function(x){
    gsub("([*+_#-])","\\\\\\1",x)
  }

# --- Escape leading Markdown list/header markers only --- 
# Markdown interprets leading "-", "+", and "#" as list/header 
# markers, which can generate unsupported HTML tags in gridtext. 
# This is required when level_labels contains "-" or "+"
  escape_leading_md <- function(x){
    gsub("^([-+#])(?=\\s|$)", "\\\\\\1", x, perl = TRUE)
  }
# --- Apply level labels while preserving intentional Markdown --- 
# Raw values not replaced by level_labels are escaped to prevent 
# unintended Markdown formatting. # Values explicitly provided through level_labels are kept unchanged 
# so intentional Markdown, e.g. "*R. solani*", remains active.
  apply_level_label <- function(x){
    if(is.null(level_labels))
      return(x)
    out <- as.character(x)

    matched <- out %in% names(level_labels)

    out[!matched] <- escape_md(out[!matched])
    out[matched]  <- level_labels[out[matched]]

    out
  }

  # --- Determine x-axis variable when faceting ---
  factors <- strsplit(as.character(unique(data$Interaction)[1]),":")[[1]]

  if (!is.null(facet_var)) {

    if (!is.null(facet_var)) {
      axis_factor <- setdiff(c("MicB", "MicP"), facet_var)

    } else {
      axis_factor <- "Interaction"
    }

    data <- data %>%
      mutate(
        .axis_value = .data[[axis_factor]]
      )

    # Apply level labels to the x-axis
    data <- data %>%
      mutate(
        .axis_value = apply_level_label(.axis_value)
      )

    # Add the corresponding treatment name
    axis_suffix <- if (!is.null(treatment_labels) &&
                      axis_factor %in% names(treatment_labels)) {
      treatment_labels[[axis_factor]]
    } else {
      ""
    }
    data <- data %>%
      mutate(
        .axis_label =
          paste0(
            .axis_value,
            ifelse(axis_suffix == "",
                  "",
                  paste0(" ", axis_suffix))
          )
      ) %>%
      mutate(.axis_label = escape_leading_md(.axis_label))
    x_aes_var <- ".axis_label"

  } else {
    x_aes_var <- "Interaction"
  }

  # --- Plot ---
  p <-
    ggplot(
      data,
      aes(
        x = .data[[x]],
        y = .data[[y]],
        colour = Interaction
      )
    )

  # --- Individual profiles ---
  if(individuals){
    p <- p +
      geom_line(
        aes(group=ID),
        alpha=individual_alpha
      )
  }
  # --- Mean profile ---
  if(average){
    p <- p +
      stat_summary(
        aes(group=Interaction),
        fun=mean,
        geom="line",
        linewidth=average_linewidth
      )
  }

  # --- Treatment colours ---
  if(!is.null(treatment_colors)){
    p <- p+
      scale_colour_manual(values=treatment_colors)
  }

  # --- Faceting ---
  if(!is.null(facet_var)){
    facet_levels_raw <- unique(as.character(data[[facet_var]]))
    facet_levels_display <- apply_level_label(facet_levels_raw)

    if(!is.null(treatment_labels) &&
       facet_var %in% names(treatment_labels)){

      facet_labeller <-
        setNames(
          escape_leading_md(
            paste0(
              facet_levels_display,
              " ",
              treatment_labels[[facet_var]]
            )
          ),
          facet_levels_raw
        )

    }else{
      facet_labeller <-
        setNames(
          escape_leading_md(facet_levels_display),
          facet_levels_raw
        )

    }
    # Apply the requested facet orientation
    if(facet_direction=="rows"){
      p <- p +
        facet_grid(
          rows=vars(.data[[facet_var]]),
          labeller=labeller(.rows=facet_labeller)
        )

    }else{
      p <- p +
        facet_grid(
          cols=vars(.data[[facet_var]]),
          labeller=labeller(.cols=facet_labeller)
        )
    }
  }

  # --- Plot style ---
  p+
    theme_bw(base_size=20)+
    theme(
      legend.position="none",

      axis.text.x = element_text(size=24),
      axis.text.y = element_text(size=24),
      axis.title  = element_text(size=26, face="bold"),

      # Override theme_bw() defaults for facet labels. 
      # Explicitly defining strip.text.x and strip.text.y ensures 
      # that Markdown is rendered correctly in both orientat
      strip.text   = ggtext::element_markdown(size=24, face="bold"),
      strip.text.x = ggtext::element_markdown(size=24, face="bold"),
      strip.text.y = ggtext::element_markdown(size=24, face="bold", angle = -90)

    )+
    labs(
      x=xlabel,
      y=ylabel
    )

}

# --- Treatment colours ---
# treatment_colors <- c(
#   "Abs:Abs" = adjust_ltc(ploen, amount = -25, which = c(3, 4))[4],
#   "Prs:Abs"    = ltc("ploen")[1], 
#   "Abs:Prs"    = adjust_ltc(ploen, amount = +25, which = c(3, 4))[5], 
#   "Prs:Prs"   = adjust_ltc(ploen, amount = -25, which = c(3, 4))[3]  
# )

treatment_colors <- c(
  "Abs:Abs" = ltc("alger")[1],
  "Prs:Abs"    = ltc("alger")[2], 
  "Abs:Prs"    = ltc("alger")[5], 
  "Prs:Prs"   = ltc("alger")[4]
)



# --- design ---
design <- read.csv("design.csv")

# --- Import and prepare bin data ---
bins <- read.csv("all_Lab_bins.csv")

bins <- bins %>%
  mutate(ID = str_extract(Image, "^\\d+"))

design <- design %>%
  mutate(ID = as.character(ID))

bins <- bins %>%
  left_join(design,
            by="ID")

bins %>%
  dplyr::select(Image, ID, MicB, MicP) %>%
  distinct()

# --- Plot all leaves ---
bins_plot <- bins %>%
  mutate(Interaction = interaction(MicB, MicP, sep = ":"))

p_L <- plot_histogram(
    data = bins_plot %>% filter(Channel == "L"),
    xlabel = "L* bin",
    facet_var = "MicP",
    treatment_colors = treatment_colors,
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )
p_a <- plot_histogram(
    data = bins_plot %>% filter(Channel == "a"),
    xlabel = "a* bin",
    facet_var = "MicP",
    treatment_colors = treatment_colors,
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )
p_b <- plot_histogram(
    data = bins_plot %>% filter(Channel == "b"),
    xlabel = "b* bin",
    facet_var = "MicP",
    treatment_colors = treatment_colors, 
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )

ggsave(
  filename = "Results/Plots/all_leaves_L.pdf",
  plot = p_L,
  width = 26,
  height = 20
)

ggsave(
  filename = "Results/Plots/all_leaves_a.pdf",
  plot = p_a,
  width = 26,
  height = 20
)

ggsave(
  filename = "Results/Plots/all_leaves_b.pdf",
  plot = p_b,
  width = 26,
  height = 20
)

# --- leafmax plots ---
bins_leafmax <- bins_plot %>%
  mutate(leaf_number = as.numeric(str_extract(Leaf,"\\d+"))) %>%
  group_by(ID) %>%
  filter(leaf_number == max(leaf_number)) %>%
  ungroup()

p_L <- plot_histogram(
    data = bins_leafmax %>% filter(Channel == "L"),
    xlabel = "L* bin",
    facet_var = "MicB",
    treatment_colors = treatment_colors,
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )
p_a <- plot_histogram(
    data = bins_leafmax %>% filter(Channel == "a"),
    xlabel = "a* bin",
    facet_var = "MicB",
    treatment_colors = treatment_colors,
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )
p_b <- plot_histogram(
    data = bins_leafmax %>% filter(Channel == "b"),
    xlabel = "b* bin",
    facet_var = "MicB",
    treatment_colors = treatment_colors, 
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )

ggsave(
  filename = "Results/Plots/6_leaves_L.pdf",
  plot = p_L,
  width = 26,
  height = 20
)

ggsave(
  filename = "Results/Plots/6_leaves_L_text.pdf",
  plot = p_L,
  width = 14,
  height = 10
)

ggsave(
  filename = "Results/Plots/6_leaves_a.pdf",
  plot = p_a,
  width = 26,
  height = 20
)


ggsave(
  filename = "Results/Plots/6_leaves_a_text.pdf",
  plot = p_L,
  width = 14,
  height = 10
)

ggsave(
  filename = "Results/Plots/6_leaves_b.pdf",
  plot = p_b,
  width = 26,
  height = 20
)

ggsave(
  filename = "Results/Plots/6_leaves_b_text.pdf",
  plot = p_L,
  width = 14,
  height = 10
)

# --- Combine L*, a*, and b* plots ---
p <- (p_a | p_b) / p_L

ggsave(
  filename = "Results/Plots/6_leaves_Lab_text.pdf",
  plot = p,
  width = 14,
  height = 16
)

# --- Average pseudo-replicates ---
bins_mean <- bins_leafmax %>%
    mutate(
    ID = as.numeric(ID),
    Replicate = ceiling(ID/2)
  ) %>%
  group_by(
    Replicate,
    Batch,
    MicB,
    MicP,
    Interaction,
    Channel,
    Bin
  ) %>%
  summarise(
    Count = mean(Count),
    .groups = "drop"
  ) %>%
  rename(ID = Replicate) %>%
  group_by(ID, Batch, MicB, MicP, Interaction, Channel) %>%
  mutate(
    Prop = Count / sum(Count)
  ) %>%
  ungroup()

p_L <- plot_histogram(
    data = bins_mean %>% filter(Channel == "L"),
    xlabel = "L* bin",
    facet_var = "MicB",
    treatment_colors = treatment_colors,
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )
p_a <- plot_histogram(
    data = bins_mean %>% filter(Channel == "a"),
    xlabel = "a* bin",
    facet_var = "MicB",
    treatment_colors = treatment_colors,
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )
p_b <- plot_histogram(
    data = bins_mean %>% filter(Channel == "b"),
    xlabel = "b* bin",
    facet_var = "MicB",
    treatment_colors = treatment_colors, 
    treatment_labels = c(MicB = "*T. asperellum*", MicP = "*R. solani*"),
    level_labels = c(Prs = "+", Abs = "-")
  )

ggsave(
  filename = "Results/Plots/6_leaves_mean_Lab_text.pdf",
  plot = p,
  width = 26,
  height = 20
)

ggsave(
  filename = "Results/Plots/6_leaves_mean_L.pdf",
  plot = p_L,
  width = 26,
  height = 20
)

ggsave(
  filename = "Results/Plots/6_leaves_mean_a.pdf",
  plot = p_a,
  width = 26,
  height = 20
)

# --- Combine L*, a*, and b* plots ---
p <- (p_a | p_b) / p_L

ggsave(
  filename = "Results/Plots/6_leaves_mean_b.pdf",
  plot = p_b,
  width = 14,
  height = 16
)
```

```{r}   
# ==========================================================
# functions
# =========================================================

# Extract the numerical leaf identifier from the "Leaf" column
extract_leaf_number <- function(data){
  data %>%
    mutate(
      leaf_number = as.numeric(
        stringr::str_extract(Leaf,"\\d+")
      )
    )
}

# link experimental design metadata to each image
add_design <- function(data, design){
  data %>%
    mutate(
      ID = stringr::str_extract(Image,"^\\d+")
    ) %>%
    left_join(
      design %>% mutate(ID=as.character(ID)),
      by="ID"
    ) %>%
    mutate(
      Interaction = paste(MicB, MicP, sep=":")
    )
}

# Select the highest-numbered leaf (here: correspond to "All segmented leaves" image) for each replicate
select_leafmax <- function(data,
                           group_vars=c("ID")){
  data %>%
    extract_leaf_number() %>%
    group_by(across(all_of(group_vars))) %>%
    filter(
      leaf_number == max(leaf_number)
    ) %>%
    ungroup()
}

# Normalize pixel histograms to relative frequencies
normalize_histogram <- function(data){
  data %>%
    group_by(
      Batch,
      Image,
      Leaf,
      Channel
    ) %>%
    mutate(
      Prop = Count / sum(Count)
    ) %>%
    ungroup()
}

# Convert 8-bit histogram bins back to original 32-bit Lab scale
#
# Reminder: Fiji macro export Lab histograms transformed in 256 bins (8-bit conversion). 
# Subsequently, orignal minimul and maximum values were recorded in Lab_ranges.csv and used to transformed  back 256 bins -> Lab values.
#
# Lab_value = (min + Bin) / (255 * (max - min))
convert_bins_to_lab <- function(data, ranges){
  ranges <- ranges %>%
    mutate(
      values = stringr::str_extract(Header,"\\[.*\\]"),
      values = stringr::str_extract_all(
        values,
        "-?[0-9]+\\.?[0-9]*"
      ),
      min = purrr::map_dbl(values, ~as.numeric(.x[1])),
      max = purrr::map_dbl(values, ~as.numeric(.x[2]))
    ) %>%
    dplyr::select(
      Batch,
      Image,
      Leaf,
      Channel,
      min,
      max
    )

  data %>%
    left_join(
      ranges,
      by=c(
        "Batch",
        "Image",
        "Leaf",
        "Channel"
      )
    ) %>%
    mutate(
      Lab_value =
        min + Bin/255*(max-min)
    )
}

# Compute weighted mean Lab values from the normalized histograms.
# Idea : Each Lab values is weighted by its pixels counts -> decrease length biais.
compute_mean_Lab <- function(data){
  data %>%
    group_by(
      ID,
      MicB,
      MicP,
      Interaction,
      Channel
    ) %>%
    summarise(
      Mean =
        weighted.mean(
          Lab_value,
          Count
        ),
      .groups="drop"
    )
}

to_CIELAB <- function(data){
  data %>%
    pivot_wider(
      names_from=Channel,
      values_from=Mean
    )
}

# Compute pairwise CIEDE2000 colo differences between treatment groups.
# Mean L*, a* and b* values are computed for each treatment.
# Then the pairwise colour differences of the mean/individual values per treatment is compute.  
compute_deltaE <- function(data,
                           group="Interaction",
                           aggregate=TRUE){
  if(aggregate){
    color <- data %>%
      group_by(.data[[group]]) %>%
      summarise(
        L=mean(L, na.rm=TRUE),
        a=mean(a, na.rm=TRUE),
        b=mean(b, na.rm=TRUE),
        .groups="drop"
      )

  } else {
    color <- data
  }

  lab <- color[, c("L", "a", "b")] %>%
    as.matrix()
  mat <- farver::compare_colour(
    from=lab,
    to=lab,
    from_space="lab",
    method="CIE2000"
  )

  if(aggregate){
    rownames(mat) <- color[[group]]
    colnames(mat) <- color[[group]]

  } else {
    rownames(mat) <- color$ID
    colnames(mat) <- color$ID
  }

  mat
}

# Halved pariwise correlation matrix
plotdE <- function(deltaE_matrix,
                    legend_title = "Delta color",
                    level_labels  = NULL,   # ex: c(Prs = "+", Abs = "-")
                    factor_labels = NULL    # ex: c("*T. asp*", "*R. sol*") - ordre positionnel, markdown ok
){
  deltaE_matrix <- as.matrix(deltaE_matrix)
  rownames(deltaE_matrix) <- make.unique(
    rownames(deltaE_matrix)
  )
  colnames(deltaE_matrix) <- make.unique(
    colnames(deltaE_matrix)
  )
  n <- nrow(deltaE_matrix)

  # --- Force the scale to start at 0 ---
  lims <- c(0, max(deltaE_matrix, na.rm = TRUE))

  df_long <- expand.grid(
    row = seq_len(n),
    col = seq_len(n)
  )
  df_long$DeltaE <- deltaE_matrix[
    cbind(
      df_long$row,
      df_long$col
    )
  ]
  # Upper triangle only
  df_long$DeltaE[
    df_long$row >= df_long$col
  ] <- NA
  df_long$From <- factor(
    rownames(deltaE_matrix)[df_long$row],
    levels = rownames(deltaE_matrix)
  )
  df_long$To <- factor(
    colnames(deltaE_matrix)[df_long$col],
    levels = colnames(deltaE_matrix)
  )
  from_levels_used <- levels(df_long$From)[-n]
  to_levels_used   <- levels(df_long$To)[-1]
  df_long <- df_long %>%
    dplyr::filter(!is.na(DeltaE))
  df_long$From <- factor(
    as.character(df_long$From),
    levels = from_levels_used
  )
  df_long$To <- factor(
    as.character(df_long$To),
    levels = to_levels_used
  )

  # --- Helper: replace each factor level and add positional prefixes ---
  apply_level_label <- function(x) {
  sapply(as.character(x), function(x) {
  # Detect and remove optional ID suffix
  has_ID <- grepl("_[0-9]+$", x)

  if (has_ID) {
    ID <- sub("^.*_([0-9]+)$", "\\1", x)
    x <- sub("_[0-9]+$", "", x)
  }
  parts <- strsplit(x, ":")[[1]]
  # Replace levels individually (Prs -> +, Abs -> -)
  if (!is.null(level_labels)) {
    parts <- ifelse(parts %in% names(level_labels), level_labels[parts], parts)
  }
  # Positional prefix
  if (!is.null(factor_labels)) {
    parts <- paste0(parts, factor_labels[seq_along(parts)],
      ifelse(is.na(factor_labels[seq_along(parts)]), "", " "))
  }
  result <- paste(parts, collapse = "  ×  ")
  # Add ID only when it exists
  if (has_ID) {
    result <- paste0(result, " [", ID, "]")
  }
  result
  })
  }

  # --- Display labels (do not replace the internal factor levels) ---
  from_labels_display <- setNames(
    apply_level_label(from_levels_used),
    from_levels_used
  )
  to_labels_display <- setNames(
    apply_level_label(to_levels_used),
    to_levels_used
  )

  ggplot(
    df_long,
    aes(
      x = From,
      y = To,
      fill = DeltaE
    )
  ) +
    geom_tile(
      color = "white",
      linewidth = 0.8
    ) +
    geom_text(
      aes(
        label = round(
          DeltaE,
          2
        )
      ),
      color = "black",
      size = 22             # --- size increase for pwp ---
    ) +
    scale_fill_gradientn(
      colours = c(
        #"white",
        ltc("alger")[2],
        ltc("alger")[4]
        
      ),
      limits = lims,
      na.value = "transparent"
    ) +
    guides(
     fill = guide_colourbar(
        title.position = "top",
        title.hjust = 0.5,
        barwidth = unit(1.5, "cm"),
        barheight = unit(12, "cm")
      )
    ) +
    scale_x_discrete(
      limits = from_levels_used,
      labels = unname(from_labels_display[from_levels_used]),   # --- Force ordering ---
      expand = c(0, 0)
    ) +
    scale_y_discrete(
      limits = rev(to_levels_used),
      labels = unname(to_labels_display[rev(to_levels_used)]),  # --- Force ordering ---
      expand = c(0, 0)
    ) +
    coord_fixed() +
    theme_bw(base_size = 20) +        # --- Base font size increased ---
    theme(
      axis.title = element_blank(),
      panel.grid = element_blank(),
      panel.border = element_blank(),
      axis.ticks = element_blank(),
      axis.text.x = ggtext::element_markdown(           # --- Italic Markdown labels ---
        angle = 45,
        hjust = 1,
        size = 48
      ),
      axis.text.y = ggtext::element_markdown(            # --- Italic Markdown labels ---
        size = 48
      ),
      legend.position = "right",
      legend.title = element_text(size = 40, face = "bold"),
      legend.text  = element_text(size = 32),
      legend.margin = margin(0, 0, 0, 0),
      legend.box.margin = margin(0, 0, 0, 0),
      plot.margin = margin(5, 5, 5, 5)
    ) +
    labs(fill = legend_title)
}

save_deltaE <- function(res,
                        plot,
                        prefix,
                        width = 26,
                        height = 20){
  ggsave(
    paste0("Results/Plots/", prefix, ".pdf"),
    plot,
    width = width,
    height = height,
    limitsize = FALSE
  )

  write.csv(
    res,
    paste0("Results/Tables/LAB_", prefix, ".csv"),
    row.names = FALSE
  )
}

# ============================================================
# Main analysis
# ============================================================
bins <- read.csv("all_Lab_bins.csv")
design <- read.csv("design.csv")
ranges <- read.csv("Lab_ranges.csv")

bins <- bins %>%
  add_design(design) %>%
  select_leafmax("ID") %>%
  normalize_histogram() %>%
  convert_bins_to_lab(ranges)

mean_Lab <- bins %>%
  compute_mean_Lab() %>%
  to_CIELAB()

# Calculate individual-image CIELAB values
Lab_ind <- bins %>%
  group_by(ID, MicB, MicP, Interaction, Channel) %>%
  summarise(
    value = weighted.mean(Lab_value, Count),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Channel,
    values_from = value
  )

Lab_ind

# --- treatment mean pairwise comparaison
deltaE <- compute_deltaE(Lab_ind, aggregate = TRUE)

#TODO : intervertir [2]:[3] ; [,2]:[,3]

new_order <- c("Abs:Abs", "Prs:Abs", "Abs:Prs", "Prs:Prs")

deltaE2 <- deltaE[new_order, new_order]

plotdE(deltaE2)

p_mean <- plotdE(deltaE2, legend_title = "delta color",  level_labels  = c(Prs = "+", Abs = "-"), factor_labels = c("*T. asp*", "*R. sol*"))

save_deltaE( Lab_ind, p_mean, "exp2_means")

# --- Individual pairwise comparaison
deltaE <- compute_deltaE(Lab_ind, aggregate = FALSE)

ids <- paste(Lab_ind$Interaction, Lab_ind$ID, sep = "_")

rownames(deltaE) <- ids
colnames(deltaE) <- ids

p_day <- plotdE(deltaE, legend_title = "delta color",  level_labels  = c(Prs = "+", Abs = "-"), factor_labels = c("*T. asp*", "*R. sol*"))

#TODO: fix Latex R_sol labels
save_deltaE( Lab_ind, p_day, "exp2_day", width = 320, height = 160)

Lab_rep <- Lab_ind %>%
  mutate(Replicate = ceiling(as.numeric(ID) / 2)) %>%
  group_by(Replicate, MicB, MicP, Interaction) %>%
  summarise(
    L = mean(L),
    a = mean(a),
    b = mean(b),
    .groups = "drop"
  ) %>%
  rename(ID = Replicate)

Lab_SD <- Lab_rep %>%
  group_by(Interaction) %>%
  summarise(
    n = n(),
    L_mean = mean(L),
    L_sd   = sd(L),

    a_mean = mean(a),
    a_sd   = sd(a),

    b_mean = mean(b),
    b_sd   = sd(b),

    .groups = "drop"
  )

Lab_SD

# ============================================================
# Generate formatted CIELAB summary tables
# ============================================================
LabTable <- function(
    df,
    out_dir = "Results",
    rep_id,
    treatment_labels = c(
      MicB = "*T.asperellum*",
      MicP = "*R.solani*"
    ),
    level_labels = c(
      Prs = "+",
      Abs = "-"
    )
){

  # --- 1. Create treatment labels 
  df <- df %>%
    tidyr::separate(
      Interaction,
      into = c("MicB", "MicP"),
      sep = ":",
      remove = FALSE
    ) %>%
    mutate(
      Treatment = paste0(
        level_labels[MicB], "",
        treatment_labels["MicB"],
        "\\ × ",
        level_labels[MicP], "",
        treatment_labels["MicP"]
      )
    )

  # --- 2. Formatting 
  df <- df %>%
    mutate(
      across(
        where(is.numeric),
        ~ signif(., 3)
      )
    ) %>%
    mutate(
      across(
        everything(),
        ~ ifelse(is.na(.) | . == "", "-", .)
      )
    )

  # --- 3. Select + rename 
  df <- df %>%
    dplyr::select(
      Treatment,
      n,
      L_mean, L_sd,
      a_mean, a_sd,
      b_mean, b_sd
    ) %>%
      rename(
        "Treatment" = Treatment,
        "n" = n,
        "L_mean" = L_mean,
        "L_SD" = L_sd,
        "a_mean" = a_mean,
        "a_SD" = a_sd,
        "b_mean" = b_mean,
        "b_SD" = b_sd
      )

  # --- 4. Table 
  tbl <- tt(
    df,
    cols = tidyselect::everything(),
    rownames = FALSE,
    escape = FALSE
  ) |>
    tt_header_above(c(
      " " = 2,
      "*L\\**" = 2,
      "*a\\**" = 2,
      "*b\\**" = 2
    )) |>
    tt_column("Treatment", width = "25%") |>
    tt_column("n", width = "6%")

  # --- 5. Export 
  dir.create(
    out_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )

  typ_file <- file.path(
    out_dir,
    paste0("Lab_", rep_id, ".typ")
  )

  pdf_file <- file.path(
    out_dir,
    paste0("Lab_", rep_id, ".pdf")
  )

  writeLines(
    capture.output(print(tbl)),
    typ_file
  )

  system(
    paste(
      "typst compile",
      typ_file,
      pdf_file
    )
  )

  return(tbl)
}

LabTable(
  Lab_SD,
  rep_id = "Lab",
  treatment_labels = c(
    MicB = "_T.asperellum_",
    MicP = "_R.solani_"
  ),
  level_labels = c(
    Prs = "+",
    Abs = "-"
  )
)
```

